The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-07 manuscript.Rmd
moar edits

Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013]. Undoubtely, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000]. Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016]. As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial. However, to understand the role of these species in their environment, we need to document the ecology of these species [@Cernansky2017]. For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements. Occurrence data for marine species could be used for environmental niche modelling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a roadmap for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richenss, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed. These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biotas. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity databases. Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the better studied phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap)). Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have not digital records for them in these geographic areas.

```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)', ech=FALSE}
knitr::include_graphics("figures/database_overlap.pdf")
```

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced datapoints and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

@@

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)).; (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.

@@

Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@xxxx]. However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx]. Based on iDigBio data, `r prop_10_specimens_species`% of species have less than 10 specimens and `r prop_30_specimens_species`% have less than 30 specimens. This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig.  \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected.

@@

Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015]. This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015]. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.

@@

The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ. It varies between 12% for bryozoans to 28% for cnidarians (\@ref(fig:proportion-barcoded)). The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available. In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.

Using the fraction of known species that have a barcode as a yard stick of the progress made is not sufficient. Many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as currently reported [@Knowlton2000,@Meyer2005-evolution]. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them). It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity. This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).

@@

Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them. Geographic information might be available in other forms, but need to be digitized. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches need to be develop for the sea.

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions. There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records. It however requires that resources are made available to museum collections in order to address these issues directly in collection databases.

Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get pooled into the large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

Another venue to improve the quality of the data in biodiversity information systems, is to develop better incitives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].

@@

Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during the preservation process. We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.

@@

Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity. We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functionning of ecosystems. In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurences for species as well as the characteristics of their habitats. Among the challenges for this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet. Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems. For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon Oil Spill [@Bik2012]. Metabarcoding has also revealed important differences in the diet of 3 related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].

Considering additional linked data with species such as DNA has been a first step in a digital biodiversity world where other data on species can be gathered in databases and analyzed to better understand species distributions. This has paved the way to more linked data/databases. Still issues though with connecting identifiers [@Guralnick2014].

Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples. Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.

Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014]. For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species. There are however opportunities for citizen science to improve biodiversity data, educate and engage the general public with marine conservation issues. For instance it has been successfully used to classify plankton images [@Robinson2017], transcribe hand-written museum labels [@Hill2012], or be involved in bioblitzes [@Ballard2017]. These initiatives however involve a lot less people compared to projects related to birds.

Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.
Digitization efforts are catching up on an already fragmented and vastly incomplete records. Here, we only reported results from the US where there is at least two centuries of biodiversity exploration data and high-level of digitization. Many other parts of the world also need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.

<!-- Biodiversity discovery and documentation has been transformed in the 21st century by molecular data to facilitate species identification, delineate species. And more recently, the development of high throughput sequencing technologies is used to quantify diversity and uncover large scale patterns of biodiversity organization. The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species, and enabled collaborative research programs such as the Census of Marine Life. Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation. Despite early criticisms and now well-understood limitations, DNA barcoding has been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling. Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available. By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification: sequences with no match make it easier to flag specimens that warrant further taxonomic attention. -->

<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult). DNA barcoding has also been seen as -->

<!-- Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding. -->

@@

<!-- here there are 3 things: data availability (having good APIs to access the
data); data comprehensiveness (how many species are captured in the database);
data accuracy (can we trust the data that is found in the databases) -->

@@

To search by geographic area, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions]. For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book. We infered a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island. For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288). These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively. The resulting occurences were then clipped to only include records that fell within the EEZ boundaries.

@@

## Supplementary material

<!--

It has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.

While plankton is recognized for its biological importance (dispersal, food source for many species), and can be used as an ecological indicator for pollution or global change, identification of planktonic larvae at the species level from morphological information is generally not possible because of the paucity of informative characters to base the identifications on, and the lack of knowledge of the appearance of the larval stages for most species. DNA barcoding can bridge this knowledge gap by using genetic information to identify larvae at the species level. These identifications are only possible if the species have already been sequenced, and the sequences deposited in public repositories. The accuracy of the identification relies on the expertise of the taxonomists who identified the specimens that have been sequenced.

We collected over the course of a year planktonic larvae of marine invertebrates in a coastal marsh of the Atlantic coast of Florida. One of our goals was to identify these larvae at the species level using DNA barcoding. We photographed and sequenced xxx larvae, representing an estimated xxx species. These data are available from BOLD and can be browsed interactively at http://plankton-atlas.net. Using records from public databases (BOLD, GenBank), we could confidently assign only xxx% to species that had already been sequenced. On average, these taxa were sequenced xxx km away from our study site.

>>>>

Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013]. Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy, and contributed to elucidating cryptic species complexes [@Knowlton2000]. Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016]. As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial. However, to understand the role of these species in their environment, we need to document the ecology of these species [@Cernansky2017]. For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements. Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the proportion of species for which we have enough data to do so is unknown.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), now allow us to evaluate what and how much we know. Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico and the Pacific Northwest, to highlight gaps in our knowledge that provide a road map for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but more comprehensive data on species distributions and abundances are needed. These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

Here, we focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories. Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

For the Gulf of Mexico, the majority of the species included in the expert faunistic list are recorded in both iDigBio and OBIS (`r format_output(species_in_common$n_spp_gom_idigbio_obis/species_in_common$n_spp_gom*100)` %), or only OBIS (`r format_output(species_in_common$n_spp_gom_obis/species_in_common$n_spp_gom*100)` %) (Fig. \@ref(fig:database-coverage)A). For the Pacific Northwest, these proportions were much lower with only `r format_output(species_in_common$n_spp_kozloff_idigbio_obis/species_in_common$n_spp_kozloff*100)`% of the species from the regional list with records in OBIS and iDigBio, and `r format_output(species_in_common$n_spp_kozloff_obis/species_in_common$n_spp_kozloff*100)`%  in OBIS only (Fig. \@ref(fig:database-coverage)B). Actually, the majority of the species in the faunistic list are not recorded in any of the public databases. Overall, there is more overlap among the data sources for the Gulf of Mexico than the Pacific Northwest.

@@

We investigated the number of species for each phylum that contributed to the discrepancies between the regional species lists and the database records. We found that the better studied phyla contributed the most to the differences across the data sources. OBIS tends to be closer to the full record set, but iDigBio has more species for the Gulf of Mexico that are not included in the list (Fig. \@ref(fig:database-overlap)). Many species of mollusks, arthropods, and annelids, known to occur in the Gulf of Mexico or in the Pacific Northwest, have no digital records in these geographic areas.

```{r compare_database_overlap, fig.cap='Comparison of species by phylum across the public databases and the faunistic list. (\\#fig:database-overlap)'}
knitr::include_graphics("figures/database_overlap.pdf")
```

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are deposited in museums less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of their range: the species were included in the list because they are known to occur within the boundaries of the geographic area, but they have never been collected within these geographic boundaries; (5) do not have geographic coordinates associated with them: some of these species can be in the database but without georeferenced data points and therefore were not returned in our queries.

Untangling the relative contribution of these factors to explain the discrepancies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepancy, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems.

@@

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century led to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and new species. Coordinated large-scale biodiversity expeditions continue to be needed to improve our understanding of the diversity of species and their spatial distributions. While less explored than the US, a recent study in Western Australia reported that the collection of 135 locations led to the collection of 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014], highlighting the potential for discovery in the oceans.

@@

Predicting the geographic range of species, and how they will be impacted by global change has undergone important advances recently [@xxxx]. However, it is generally recognized that to obtain reliable results such efforts require to have accurate geographic information for a minimum of 10 specimens but having at least 30 is preferred [@xxxx]. From the iDigBio data, we found that `r prop_10_specimens_species`% of species are known from less than 10 records and `r prop_30_specimens_species`% from less than 30 records. This suggests that our understanding of the geographic range of species is limited as most species are known from only a few locations, while efforts to estimate their current and future ranges based on occurrence data are rendered moot by the low number of data points available.

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens collected recently remain for the most part unidentified in collections (Fig. \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but taxonomic experts continue to be needed to identify specimens recently collected in the US.

@@

Since its introduction in the early 2000s, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples cost effectively [@Leray2015]. This approach opens up new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by orders of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching DNA sequences obtained to previously sequenced and correctly identified species [@Leray2015]. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases.

@@

The proportion of species that have DNA barcodes available is relatively constant across the better studied phyla at the scale of the EEZ. It varies between 12% for bryozoans to 28% for cnidarians (Fig. @ref(fig:proportion-barcoded)). The proportions for species known to occur in the Gulf of Mexico are usually similar than for the US, while typically a higher fraction of species from the Pacific Northwest are available. In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g. 2.6% for the Antarctic [@Grant2009]). This suggests that even as many additional species have had their DNA barcodes made available since these studies, the US fauna is generally better characterized than other parts of the world.

Using the fraction of known species that have a barcode as a yard stick of the progress made is not sufficient. Many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as currently reported [@Knowlton2000,@Meyer2005-evolution]. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. Geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records, big.mark=",")` BOLD records analyzed here have no geographic coordinate information associated with them). It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that enough samples for each species are sequenced, to capture more accurately their geographic range and the extent of their genetic diversity. This cryptic diversity is also at least partially reflected in the fact that `r stat_bold$n_spp_multi_bin` species out of `r format(stat_bold$n_spp, big.mark=",")` in our dataset include more than 1 BIN (BOLD Index Number, operational taxonomic units that closely resemble species [@Ratnasingham2013]).

@@

Similarly, a significant fraction of museum records have missing or inaccurate geographic information associated with them. Geographic information might be available in other forms, but needs to be digitized. While tools have been developed to convert names of locales into GPS coordinates on land [@rios2010geolocate], similar approaches are lacking for the sea.

For both the spelling of scientific names, and the geographic information, computing tools can be developed to detect issues and propose solutions. There are opportunities to develop computing services that use publicly available data to flag and fix these issues directly from collection databases records (e.g. see [@Otegui2016] for vertebrate records). It however requires that resources are made available to museum collections in order to address these issues directly in collection databases.

Increasing the computing literacy and teaching data wrangling skills to staff of museums in order for them to manipulate and use the data from their collection database is equally critical. As data across institutions get incorporated and pooled into large biodiversity information systems, it is crucial that data standards such as Darwin Core are embraced by the community.

Another venue to improve the quality of the data in biodiversity information systems, is to develop better incentives from journal editors and funding agencies for scientists to publish their occurrence records in databases alongside their papers as it is done for genetic data with GenBank [@Huang2011e].

@@

Samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during the preservation process. We also need to capture metadata associated with specimens to be able to understand better the physical characteristics of the habitats and the life history for these species. Specimens need to be sub-sampled to create material suitable for genomic analysis. Parts of data acquisition associated with specimens and their biota can be automated and captured digitally directly in the field, speeding up their ingestion in database collections.

@@

Assembling species lists is a necessary but not sufficient step in our quest to understand biodiversity. We also need to understand better the geographic distribution of biodiversity, how the abundance of species vary in space and time, and their role in the functioning of ecosystems. In order to predict the impact of climate change and other anthropogenic perturbations on species distributions, we need to collect additional occurrences for species as well as the characteristics of their habitats. Among the challenges of this enterprise, we need to be able to convey taxonomic concepts for species which, for the most part, have not formally been named yet. Developing robust identifiers that can be shared across biodiversity information systems, and anchored in results from DNA barcoding or other molecular techniques, is needed to collaboratively build our knowledge of species distributions.

Metabarcoding also offers to be a promising technology to monitor changes in communities and for understanding the functioning of ecosystems. For instance, despite our limited knowledge of the diversity of communities in the meiofauna, metabarcoding has been successfully used to detect and documents changes in species assemblages following the Deepwater Horizon oil spill [@Bik2012]. Metabarcoding has also revealed important differences in the diet of three related species of fishes that co-occur at small spatial scales, providing a much more detailed understanding of trophic relations [@Leray2015b].

## Conclusion

Given the urgency of the biodiversity crisis, we need to develop data-driven initiatives to efficiently document biodiversity, and build on our existing knowledge to collect information to improve our understanding of marine systems.

DNA sequences have been one of the first test case on how to link biodiversity data across repositories. These initial efforts have revealed the lack of infrastructure to use data formats designed for linked data, the lack of ontologies, and the fragility of the identifiers used [@Stucky2014; @Walls2014; @Guralnick2014]. As new types of data will be added and linked to existing repositories, the community needs to develop modern and robust ways to integrate heterogeneous biodiversity data sources [@Guralnick2015; @Deck2015].

Digitization efforts are catching up on already fragmented and incomplete records. Even if technological advances can greatly speed up the documentation of species and their distributions, taxonomists and experts are needed  to collect, sort, process, and identify samples. Taxonomic expertise within the US is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections.

Crowd-sourced data and citizen science initiatives have been proposed to augment data about species abundance and distributions [@Pimm2014]. For marine taxa, beyond the application to near-shore habitats on easily recognizable species [@Delaney2008], large-scale efforts are currently not possible given the specialized equipment needed to sample most marine habitats and the expertise required to identify species. There are however opportunities for citizen science to help curate biodiversity data, while educating and engaging the public with marine conservation issues. For instance it has been successfully used to classify plankton images [@Robinson2017], or transcribe hand-written museum labels [@Hill2012].

Here, we reported results from the US where there is at least two centuries of biodiversity exploration data and high-levels of digitization of museum collections. Many other parts of the world need additional attention, especially in some developing countries where environmental pressure on ecosystems is even higher than in the US.

@@

<!-- here there are 4 things: data availability (having good APIs to access the
data); data comprehensiveness (how many species are captured in the database);
data accuracy (can we trust the data that is found in the databases); data linkage -->

@@

To search by geographic area, we obtained the boundaries of the Exclusive Economic Zone for the United States from marineregions.org (mrgid:8456) [@mregions]. For the Pacific Northwest, Kozloff did not specify the exact geographic extent covered by his book. We inferred a polygon that encompass part of the Oregon coast to the northern parts of Vancouver island. For the Gulf of Mexico, we used the same geographic area as the one defined in [@Tunnell2009] by modifying manually the eastern boundaries of the shape file for the Gulf of Mexico provided by marineregions.org (mrgid:4288). These geographic areas were split into squares of 0.2°x0.2° which were used to query the web API of iDigBio and OBIS using the R packages ridigbio and robis respectively. The resulting occurrences were then clipped to only include records that fell within the EEZ boundaries.

@@

<!-- ## Supplementary material -->
